Decisions Decisions Decisions: Carrying out Latent Profile Analysis in Accordance With Best Practices Using the tidyLPA R package

### 2020-12-29 paper.rmd
format paper

> Grades are normally distributed, with $\mu$ = 75, $\sigma$ = 5.

@@

```{r, include = FALSE, eval = TRUE}
def <- defData(varname = "x1", formula = 65, variance = 30)
def <- defData(def, varname = "x2", formula = 85, variance = 23)
def <- defData(def, varname = "xMix", 
               formula = "x1 | .5 + x2 | .5", 
               dist = "mixture")

set.seed(2716)
dx <- genData(1000, def)
```

@@

```{r, include = FALSE, eval = TRUE}
ggplot(dx, aes(x = xMix)) +
  geom_density() +
  papaja::theme_apa() +
  ylim(0, .10)
```

@@

```{r, include = FALSE, eval = TRUE}
dx %>% 
  gather(key, val, x1:x2) %>% 
  ggplot(aes(x = val, fill = key)) +
  geom_density(alpha = .3) +
  papaja::theme_apa() +
  ylim(0, .10)
```

@@

```{r, echo = FALSE}
t <- readr::read_csv("GRoLTS-applicability-to-LPA.csv")

tp <- t %>% 
  filter(use == 1) %>% 
  mutate(`LPA Number` = 1:nrow(.)) %>% 
  select(`GRoLTS Number` = Number, `LPA Number`, `Checklist Item`)

tp %>% 
  DT::datatable()
```

@@

```{r}
pisaUSA15 %>%
  select(broad_interest, enjoyment, self_efficacy) %>% 
  tidyr::gather(key, val) %>% 
  ggplot(aes(x = val, group = key, fill = key)) + 
  geom_density() + 
  facet_wrap(~key) + 
  theme_bw()
```

@@

```{r}
pisaUSA15 %>%
  select(broad_interest, enjoyment, self_efficacy) %>% 
  psych::describe() %>% 
  knitr::kable()
```

@@

We can see a number of statistics for each of the models that were estimated, including the 
AIC (Aikake information criterion; based on -2 * the log-likelihood, and penalized by number of parameters) and BIC (BIC: Bayesian information criterion; based on -2 log-likelihood, and penalized by number of parameters adjusted by sample size). Both are _penalized_ likelihood statistics; the log-likelihood will always decrease with the addition of additional 
paremeters estimated by the model, whereas the AIC and BIC will account for the number of parameters; lower values of
either indicate better fit.

@@

```{r, get-fit-m1}
pisaUSA15 %>%
  select(broad_interest, enjoyment, self_efficacy) %>%
  single_imputation() %>%
  estimate_profiles(3) %>% 
  get_fit(m1)
```

@@

````{r, compare-mclust, eval= TRUE}
many_analyses <- pisaUSA15 %>%
  select(broad_interest, enjoyment, self_efficacy) %>%
  single_imputation() %>%
  estimate_profiles(models = 1,
                    n_profiles = 1:6)

many_analyses

compare_solutions(many_analyses)

many_analyses_1_3 <- pisaUSA15 %>%
  select(broad_interest, enjoyment, self_efficacy) %>%
  single_imputation() %>%
  estimate_profiles(models = c(1, 3),
                    n_profiles = 1:6)

many_analyses_1_3

compare_solutions(many_analyses_1_3)
```

@@

```{r}
models_4_5 <- pisaUSA15[1:100, ] %>%
  select(broad_interest, enjoyment, self_efficacy) %>%
  single_imputation() %>%
  estimate_profiles(models = 1,
                    n_profiles = c(4, 5))

models_4_5 %>% 
  plot_profiles(add_line = TRUE)
```

@@

```{r}
models_4_5 %>% 
  plot_profiles(add_line = TRUE)
```

@@

```{r}
models_4_5 %>% 
  plot_bivariate()
```

```{r, eval = FALSE, echo = FALSE}
many_analyses_mplus <- pisaUSA15 %>%
  select(broad_interest, enjoyment, self_efficacy) %>%
  single_imputation() %>%
  estimate_profiles(models = c(1),
                    n_profiles = 1:7,
                    package = "mplus")

many_analyses_mplus

compare_solutions(many_analyses_mplus)
```

### Specifying the _model_

@@

### 1. Equal variances, and covariances fixed to 0 (model 1)

@@

$$
\left[ \begin{matrix} { \sigma  }_{ 1 }^{ 2 } & 0 & 0 & 0 \\ 0 & { \sigma  }_{ 2 }^{ 2 } & 0 & 0 \\ 0 & 0 & { \sigma  }_{ 3 }^{ 2 } & 0 \\ 0 & 0 & 0 & { \sigma  }_{ 4 }^{ 2 } \end{matrix} \right] 
$$

### 2. Varying variances and covariances fixed to 0 (model 2)

@@

$$ 
\left[ \begin{matrix} { \sigma  }_{ 1p }^{ 2 } & 0 & 0 & 0 \\ 0 & { \sigma  }_{ 2p }^{ 2 } & 0 & 0 \\ 0 & 0 & { \sigma  }_{ 3p }^{ 2 } & 0 \\ 0 & 0 & 0 & { \sigma  }_{ 4p }^{ 2 } \end{matrix} \right] 
$$

### 3. Equal variances and equal covariances (model 3)

@@

$$
\left[ \begin{matrix} { \sigma  }_{ 1 }^{ 2 } & { \sigma  }_{ 21 } & { \sigma  }_{ 31 } & { \sigma  }_{ 41 } \\ { \sigma  }_{ 12 } & { \sigma  }_{ 2 }^{ 2 } & { \sigma  }_{ 23 } & { \sigma  }_{ 24 } \\ { \sigma  }_{ 13 } & { \sigma  }_{ 12 } & { \sigma  }_{ 3 }^{ 2 } & { \sigma  }_{ 33 } \\ { \sigma  }_{ 14 } & { \sigma  }_{ 12 } & { \sigma  }_{ 12 } & { \sigma  }_{ 4 }^{ 2 } \end{matrix} \right] 
$$

### 4. Varying means, varying variances, and equal covariances (model 4)

@@

$$
\left[ \begin{matrix} { \sigma  }_{ 1p }^{ 2 } & { \sigma  }_{ 21 } & { \sigma  }_{ 31 } & { \sigma  }_{ 41 } \\ { \sigma  }_{ 12 } & { \sigma  }_{ 2p }^{ 2 } & { \sigma  }_{ 23 } & { \sigma  }_{ 24 } \\ { \sigma  }_{ 13 } & { \sigma  }_{ 12 } & { \sigma  }_{ 3p }^{ 2 } & { \sigma  }_{ 33 } \\ { \sigma  }_{ 14 } & { \sigma  }_{ 12 } & { \sigma  }_{ 12 } & { \sigma  }_{ 4p }^{ 2 } \end{matrix} \right] 
$$

### 5. Varying means, equal variances, and varying covariances (model 5)

@@

$$
\left[ \begin{matrix} { \sigma  }_{ 1 }^{ 2 } & { \sigma  }_{ 21p } & { \sigma  }_{ 31p } & { \sigma  }_{ 41p } \\ { \sigma  }_{ 12p } & { \sigma  }_{ 2 }^{ 2 } & { \sigma  }_{ 23p } & { \sigma  }_{ 24p } \\ { \sigma  }_{ 13p } & { \sigma  }_{ 12p } & { \sigma  }_{ 3 }^{ 2 } & { \sigma  }_{ 33p } \\ { \sigma  }_{ 14p } & { \sigma  }_{ 12p } & { \sigma  }_{ 12p } & { \sigma  }_{ 4 }^{ 2 } \end{matrix} \right] \quad 
$$

### 6. Varying variances and varying covariances (model 6)

@@

$$
\left[ \begin{matrix} { \sigma  }_{ 1p }^{ 2 } & { \sigma  }_{ 21p } & { \sigma  }_{ 31p } & { \sigma  }_{ 41p } \\ { \sigma  }_{ 12p } & { \sigma  }_{ 2p }^{ 2 } & { \sigma  }_{ 23p } & { \sigma  }_{ 24p } \\ { \sigma  }_{ 13p } & { \sigma  }_{ 12p } & { \sigma  }_{ 3p }^{ 2 } & { \sigma  }_{ 33p } \\ { \sigma  }_{ 14p } & { \sigma  }_{ 12p } & { \sigma  }_{ 12p } & { \sigma  }_{ 4p }^{ 2 } \end{matrix} \right] 
$$

>>>>

> Grades are normally distributed, with $\\mu$ = 75, $\\sigma$ = 5.

@@

``` {r, include = FALSE, eval = TRUE}
def <- defData(varname = "x1", formula = 65, variance = 30)
def <- defData(def, varname = "x2", formula = 85, variance = 23)
def <- defData(def, varname = "xMix", 
               formula = "x1 | .5 + x2 | .5", 
               dist = "mixture")
```

set.seed(2716) dx \<- genData(1000, def)

``` 
```

@@

``` {r, include = FALSE, eval = TRUE}
ggplot(dx, aes(x = xMix)) +
  geom_density() +
  papaja::theme_apa() +
  ylim(0, .10)
```

@@

``` {r, include = FALSE, eval = TRUE}
dx %>% 
  gather(key, val, x1:x2) %>% 
  ggplot(aes(x = val, fill = key)) +
  geom_density(alpha = .3) +
  papaja::theme_apa() +
  ylim(0, .10)
```

@@

``` {r, echo = FALSE}
t <- readr::read_csv("GRoLTS-applicability-to-LPA.csv")
```

tp \<- t %\>% filter(use == 1) %\>% mutate(`LPA Number` = 1:nrow(.)) %\>%
select(`GRoLTS Number` = Number, `LPA Number`, `Checklist Item`)

tp %\>% DT::datatable()

``` 
```

@@

``` {r}
pisaUSA15 %>%
  select(broad_interest, enjoyment, self_efficacy) %>% 
  tidyr::gather(key, val) %>% 
  ggplot(aes(x = val, group = key, fill = key)) + 
  geom_density() + 
  facet_wrap(~key) + 
  theme_bw()
```

@@

``` {r}
pisaUSA15 %>%
  select(broad_interest, enjoyment, self_efficacy) %>% 
  psych::describe() %>% 
  knitr::kable()
```

@@

We can see a number of statistics for each of the models that were estimated,
including the AIC (Aikake information criterion; based on -2 \* the
log-likelihood, and penalized by number of parameters) and BIC (BIC: Bayesian
information criterion; based on -2 log-likelihood, and penalized by number of
parameters adjusted by sample size). Both are *penalized* likelihood statistics;
the log-likelihood will always decrease with the addition of additional
paremeters estimated by the model, whereas the AIC and BIC will account for the
number of parameters; lower values of either indicate better fit.

@@

``` {r, get-fit-m1}
pisaUSA15 %>%
  select(broad_interest, enjoyment, self_efficacy) %>%
  single_imputation() %>%
  estimate_profiles(3) %>% 
  get_fit(m1)
```

@@

``` {r, compare-mclust, eval= TRUE}
many_analyses <- pisaUSA15 %>%
  select(broad_interest, enjoyment, self_efficacy) %>%
  single_imputation() %>%
  estimate_profiles(models = 1,
                    n_profiles = 1:6)
```

many\_analyses

compare\_solutions(many\_analyses)

many\_analyses\_1\_3 \<- pisaUSA15 %\>% select(broad\_interest, enjoyment,
self\_efficacy) %\>% single\_imputation() %\>% estimate\_profiles(models =
c(1, 3), n\_profiles = 1:6)

many\_analyses\_1\_3

compare\_solutions(many\_analyses\_1\_3)

``` 
```

@@

``` {r}
models_4_5 <- pisaUSA15[1:100, ] %>%
  select(broad_interest, enjoyment, self_efficacy) %>%
  single_imputation() %>%
  estimate_profiles(models = 1,
                    n_profiles = c(4, 5))
```

models\_4\_5 %\>% plot\_profiles(add\_line = TRUE)

``` 
```

@@

``` {r}
models_4_5 %>% 
  plot_profiles(add_line = TRUE)
```

@@

``` {r}
models_4_5 %>% 
  plot_bivariate()
```

``` {r, eval = FALSE, echo = FALSE}
many_analyses_mplus <- pisaUSA15 %>%
  select(broad_interest, enjoyment, self_efficacy) %>%
  single_imputation() %>%
  estimate_profiles(models = c(1),
                    n_profiles = 1:7,
                    package = "mplus")
```

many\_analyses\_mplus

compare\_solutions(many\_analyses\_mplus)

``` 
```

### Specifying the *model*

@@

### 1\. Equal variances, and covariances fixed to 0 (model 1)

@@

$$ \\left[ \\begin{matrix} { \\sigma }*{ 1 }^{ 2 } & 0 & 0 & 0 \\ 0 & { \\sigma
}*{ 2 }^{ 2 } & 0 & 0 \\ 0 & 0 & { \\sigma }*{ 3 }^{ 2 } & 0 \\ 0 & 0 & 0 & {
\\sigma }*{ 4 }^{ 2 } \\end{matrix} \\right] $$

### 2\. Varying variances and covariances fixed to 0 (model 2)

@@

$$ \\left[ \\begin{matrix} { \\sigma }*{ 1p }^{ 2 } & 0 & 0 & 0 \\ 0 & {
\\sigma }*{ 2p }^{ 2 } & 0 & 0 \\ 0 & 0 & { \\sigma }*{ 3p }^{ 2 } & 0 \\ 0 & 0
& 0 & { \\sigma }*{ 4p }^{ 2 } \\end{matrix} \\right] $$

### 3\. Equal variances and equal covariances (model 3)

@@

$$ \\left[ \\begin{matrix} { \\sigma }*{ 1 }^{ 2 } & { \\sigma }*{ 21 } & {
\\sigma }*{ 31 } & { \\sigma }*{ 41 } \\ { \\sigma }*{ 12 } & { \\sigma }*{ 2
}^{ 2 } & { \\sigma }*{ 23 } & { \\sigma }*{ 24 } \\ { \\sigma }*{ 13 } & {
\\sigma }*{ 12 } & { \\sigma }*{ 3 }^{ 2 } & { \\sigma }*{ 33 } \\ { \\sigma
}*{ 14 } & { \\sigma }*{ 12 } & { \\sigma }*{ 12 } & { \\sigma }*{ 4 }^{ 2 }
\\end{matrix} \\right] $$

### 4\. Varying means, varying variances, and equal covariances (model 4)

@@

$$ \\left[ \\begin{matrix} { \\sigma }*{ 1p }^{ 2 } & { \\sigma }*{ 21 } & {
\\sigma }*{ 31 } & { \\sigma }*{ 41 } \\ { \\sigma }*{ 12 } & { \\sigma }*{ 2p
}^{ 2 } & { \\sigma }*{ 23 } & { \\sigma }*{ 24 } \\ { \\sigma }*{ 13 } & {
\\sigma }*{ 12 } & { \\sigma }*{ 3p }^{ 2 } & { \\sigma }*{ 33 } \\ { \\sigma
}*{ 14 } & { \\sigma }*{ 12 } & { \\sigma }*{ 12 } & { \\sigma }*{ 4p }^{ 2 }
\\end{matrix} \\right] $$

### 5\. Varying means, equal variances, and varying covariances (model 5)

@@

$$ \\left[ \\begin{matrix} { \\sigma }*{ 1 }^{ 2 } & { \\sigma }*{ 21p } & {
\\sigma }*{ 31p } & { \\sigma }*{ 41p } \\ { \\sigma }*{ 12p } & { \\sigma }*{ 2
}^{ 2 } & { \\sigma }*{ 23p } & { \\sigma }*{ 24p } \\ { \\sigma }*{ 13p } & {
\\sigma }*{ 12p } & { \\sigma }*{ 3 }^{ 2 } & { \\sigma }*{ 33p } \\ { \\sigma
}*{ 14p } & { \\sigma }*{ 12p } & { \\sigma }*{ 12p } & { \\sigma }*{ 4 }^{ 2 }
\\end{matrix} \\right] \\quad $$

### 6\. Varying variances and varying covariances (model 6)

@@

$$ \\left[ \\begin{matrix} { \\sigma }*{ 1p }^{ 2 } & { \\sigma }*{ 21p } & {
\\sigma }*{ 31p } & { \\sigma }*{ 41p } \\ { \\sigma }*{ 12p } & { \\sigma
}*{ 2p }^{ 2 } & { \\sigma }*{ 23p } & { \\sigma }*{ 24p } \\ { \\sigma }*{ 13p
} & { \\sigma }*{ 12p } & { \\sigma }*{ 3p }^{ 2 } & { \\sigma }*{ 33p } \\ {
\\sigma }*{ 14p } & { \\sigma }*{ 12p } & { \\sigma }*{ 12p } & { \\sigma }*{ 4p
}^{ 2 } \\end{matrix} \\right] $$
